Thesis: draft

### 2021-05-12 draft.tex
made small changes

On-line social platforms use algorithms to order and filter the posts that appear on the time-line of an individual \cite{Perra2019}. They do this in order to make the experience on their platforms more pleasant, interesting and convenient \cite{Perra2019}. If social media companies would not use any kind of filtering, there would be way to many content to handle and process for the social media user \cite{Bozdag2013}. This in turn would make the on-line social experience unpleasant and inconvenient; the user might feel overwhelmed by the number of information he/she is exposed to \cite{Bozdag2013}. The reason behind this is that people are cognitive and time constrained \cite{Perra2019}; hence, there is only a limited amount of information that humans can process \cite{Bozdag2013}.\\
\newline
The downside of such algorithmic filtering is that it fine-tunes the information it shows based on the user. Even two users with the same friends may then be exposed to different content, based on their previous activities and history \cite{Bozdag2013}. This may lead to the formation of echo chambers in which users get `trapped' in content that confirms their own mindset \cite{Bozdag2013}. \\
\newline
It is important to know and be aware of the possible impacts of algorithmic personalization on the formation and evolution of opinions. Perra and Rocha already partly investigated the role of such algorithms on the dynamics of opinions in several networks in one of their previous papers, \textit{``Modelling opinion dynamics in the age of algorithmic personalization''} \cite{Perra2019}. Since filtering algorithms are often not transparently shared by social media companies, they proposed several personalization algorithms based on three simple ways of filtering \cite{Perra2019}: (i) Popularity filtering: promoting content that is popular across the platform; (ii) Semantic filtering: promoting posts that are similar to previous consumed posts; (iii) Collaborative filtering: promoting content that people similar or close to us consume.\\ 
\newline
The following paragraph briefly reviews the filtering algorithms that Perra and Rocha proposed. \\
\newline
At every time $t$ each platform user has a hidden list $L_i(t)$ that contains all the opinions/posts that he/she received from his/her friends \cite{Perra2019}. The filtering algorithm then selects $S$ items from this list and shows them on the user's time-line (this is the fundamental idea behind algorithmic personalization: only showing a fraction of the possible posts \cite{Perra2019}); the following selection methods are used \cite{Perra2019}: (i) The reference filtering algorithm (REF): this filtering algorithm randomly selects $S$ items from the list $L_i(t)$; (ii) The old filtering algorithm (OLD): the $S$ oldest post in $L_i(t)$ are displayed on the time-line of the user; (iii) The recent filtering algorithm (REC): shows the $S$ most recent items from $L_i(t)$; (iv) The preference filtering algorithm (PR): sorts the hidden list $L_i(t)$ according to the current opinion of the user and then selects the first $S$ elements. This filtering algorithm is based on the semantic filtering explained above.\\
\newline
Perra and Rocha showed that the REF, OLD and REC filtering algorithms behave similar in many cases and that the PR filtering algorithm often displays different behavior (such as a larger formation of echo chambers compared to the other three filtering algorithms) \cite{Perra2019}. In this thesis the REC and REF filtering algorithms are mainly used as a reference/benchmark algorithm and the PR filtering algorithm is used as the main filtering algorithm.

@@

Most of the simple opinion dynamics models make use of a static network in which users are always active and always exchanging opinions/posts with each other. In this kind of model the opinion dynamics might go as follows: each time step a random user is selected after which his/her opinion is updated according to the opinion dynamics model that is implemented. Another possibility is updating all users at the same time during each time step. \\
\newline
It might, however, be a more realistic situation if nodes/users become active/inactive over time. This would be a better representation of real social media platforms, where users are active for some periods of time and then become inactive again. This kind of activation mechanism of the users gives the network a more temporal character. The underlying network structure does not change (in the sense that each user/node has a fixed number of neighbors that does not change over time, so no new friends are made or, conversely, no old friends are being defriended), but by making users active and inactive the exchange of information between users becomes time dependent and depends on which users are active and which are not. Hence, this does affect the network structure in some way, because if a user, for example, does not become active for a long period of time, he/she does not participate in the exchange of information/opinions and the network structure behaves like if this user/node is absent. This may affect the dynamics, since it may seem as if the network has a different average degree and some properties like connectivity, clustering coefficient, etc may behave differently. \\
\newline
In this thesis we follow the same activation mechanism as was proposed in the paper of Perra and Rocha \cite{Perra2019}. Every time step users are made active with some probability $p_i$ (thus, every time step, each node becomes active with a probability $p_i$). In this thesis the activation probability is set equal to $p = 0.1$, which is constant for all the nodes. Perra and Rocha chose, in their paper, activation probabilities that are extracted from a power-law ($F(p) \sim p^{-1.5}$, with $p \in \interval{0.01}{1}$) \cite{Perra2019}. They did this based on the analyses of real social networks that shows that the propensity of users to start an on-line social interaction follows a heterogeneous distribution \cite{Perra2012}\cite{Perra2019}. However, they also showed that using a constant activation probability $p$ over the entire population yields results that are qualitatively similar (at least in their case) \cite{Perra2019}. \\
\newline
For simplicity, this thesis implements the constant activation probability variant. It might however be an interesting extension to switch to the heterogeneous activation probabilities and investigate the possible impacts on the dynamics. \\
\newline
If a user is active, he/she first spreads his current opinion to all his/her neighbors. Then, the active user updates his own opinion according to the opinions on his/her time-line \cite{Perra2019}. This updating goes according to the implemented opinion dynamics model (eg. majority model, probabilistic majority model,...). In every time step the updating of all active nodes is done simultaneously \cite{Perra2019}.

@@

It is well known that in real life, you can find people that are persistent towards their own opinion and that are not open to opinions of others. They are convinced that they carry the truth with them and that other people, with different world-views, are not capable of giving them any new and useful insights. They are close-minded in contrast to people with an open-mind that are willing to learn from others and to reconsider their beliefs. Of course this is put a bit simplistic and there are many shades of gray between this black and white statement. However, there are groups in society, often with more extreme opinions, that are know to be really inflexible towards changing that (extreme) opinion (note that being stubborn and being extreme does not need to go hand in hand, you can be stubborn without being extreme and vice versa; however, it is often widely accepted that there is some kind of correlation/link between people with extreme opinions and being more persistent towards these opinions compared to people with more moderate beliefs; this idea of correlating extreme people with a more stubborn character is also often used in models, see for example \cite{Deffuant2006}\cite{Deffuant2002}\cite{Turner}) (also note that having some resistance to change your opinion to any other opinion does not necessarily need to be a bad thing, it can be seen as a form of critical thinking and as a form of taking things with a grain of salt).\\ 
\newline
Knowing the impact of stubborn groups on the opinions of others is of great importance, especially with the rise of social media which makes the reach of more extreme and persistent groups a lot bigger. Before the era of social media, one could maybe argue that these extreme and stubborn groups are often isolated and therefore would not reach too many people. So their impact could only be of a minor nature (this is only a statement that could have been given, there are of course plenty of examples throughout human history that may contradict this argument; to only mention a few: Nazism in the 1930s and 1940s, the Ku Klux Klan in the late 19th and begin 20th century, the rise of communism in the 20th century, etc, all examples of groups that can be considered extreme and that have had a (relatively) wide reach and a significant impact on society). Today, this possible argument is no longer valid; social media enables the contact with all sorts of beliefs and all sorts of people with almost no (time and spatial) constraints. Extreme and stubborn groups that may have been isolated before no longer need to be so and this may have huge, and largely unknown, impacts. \\
\newline
What is the impact of having stubborn/extreme people in your society? Does it matter if the stubborn people are grouped together or if they are distributed evenly throughout your system? Are there ways to reduce the possible impacts of stubbornness and/or extremism? Can these stubborn groups lead to polarization or even overtake the opinions in the system? Or is introducing some kind of stubbornness, on the other hand, a protective factor towards extremism/polarization? To answer these questions, several models have been proposed to include some kind of stubbornness/resistance in the system under study.

@@

The bounded confidence models, discussed in Section \ref{bounded}, can be easily extended to included some notion of stubbornness. Guillaume Deffuant proposed such an extension where stubbornness is correlated to extremism \cite{Deffuant2006}. The model he proposed goes as follows. Remember that in the bounded confidence models, the agents have an opinion $x_i$ (which is a real number between $\interval{-1}{1}$) and a threshold $\epsilon_i$, which determines the range of opinions that can influence agent $i$ (agents with opinions between $\interval{x_i - \epsilon_i}{x_i + \epsilon_i}$ are capable of influencing agent $i$). Guillaume Deffuant then introduced a fraction of extremists, $f_e$, which can have either the extreme opinion $-1$ or the extreme opinion $1$ \cite{Deffuant2006}. These extremists are given a threshold equal to $\epsilon_e$ \cite{Deffuant2006}. The other agents are considered to be moderate and can have opinions between $-1$ and $1$. The moderates are given a threshold $\epsilon$, with $\epsilon > \epsilon_e$ \cite{Deffuant2006}. This last condition implicitly denotes the resistance of the extremes towards other opinions in contrast to the more flexible nature of the moderates, which is based on the idea that people with extreme opinions will be more convinced of that opinion and will be less likely to change it.\\
\newline
Since this thesis is not concerned with bounded confidence models, we will not discuss the results of this model here. Interested readers can find them in reference \cite{Deffuant2006}. \\
\newline
Another example of introducing some kind of persistence towards the own opinion of the nodes is the ``hardening of positions'' that was briefly discussed at the end of Section \ref{classMod}. The model introduced by Friedkin and Johnsen in which each agent adheres to its initial opinion to a certain degree, see Section \ref{classMod}, can also be seen as introducing a notion of stubbornness \cite{Noorazar2020}.

@@

One small remark must be made here: previously, the link between stubbornness and extremism was often made and it was often assumed that the more extreme somebodies opinions are, the more persistent that somebody is towards those opinions. In binary models, however, there are only two possible opinions that can be taken on and in that sense it is less useful to talk about extreme and moderate opinions. However, we can still have people that rather keep their own opinion than be convinced by friends to switch to the other opinion.\\

@@

There are different possibilities of implementing this kind of model. One way, which makes use of the majority model, is as follows. First the fraction of friends with opinion A and opinion B, $p_\text{A}$ and $p_\text{B}$ respectively, are determined; then the following rules are used to update the opinion: if $p_\text{A} > p_\text{B}$ and $p_\text{A} > T$, opinion A is adopted; if $p_\text{B} > p_\text{A}$ and $p_\text{B} > T$, opinion B is adopted; in all other cases, the current opinion of the node is not changed.\\

@@

If $T = 0$, this model reduces to the pure majority model described in Section \ref{majModel}. If $T = 1$, the nodes do not change their opinion. For $0 < T < 1$, the model is a majority model, but the nodes are a bit more persistent towards their own opinion and need a larger fraction of friends to convince them to switch to the other viewpoint.\\

@@

Note that the above examples are only a handful of the possible ways to include some notion of stubbornness in the model, many more exist or can be invented.

@@

The previous two chapters, Chapter \ref{chap2} and \ref{chap3}, gave a broad introduction/overview of networks and opinion dynamics models respectively. In this chapter we will summarize the specific networks and opinion dynamics models that are used in this thesis. \\
\newline
Section \ref{networkmodel} gives an overview of the network models that are used. In chapter \ref{chap2}, Section \ref{netModel} we discussed their main properties; here we will discuss why we use these networks. Which of their properties are of interest in this thesis? What are their benefits and possible shortcomings regarding to our goals?\\
\newline
In section \ref{realnetwork}, we will discuss the used real-world networks. The reasons of choosing these specific networks will be explained, together with their possible strengths and weaknesses.\\ % + explain where they come from etc?
\newline
Finally, Section \ref{opinionmodel} summarizes and discusses the opinion dynamics model that is implemented in this thesis. Again we will give some focus on possible advantages and disadvantages of the model.

@@

This thesis uses three main network models: (i) The Erd\H{o}s-R\'{e}nyi model (ER); (ii) The Watts-Strogatz model (WS); (iii) The stochastic block model, of which two types are implemented: the regular stochastic block model (SBM) and the stochastic Watts-Strogatz block model (SBM-WS).\\

@@

Each of these networks serve specific goals which will shortly be discussed in the following paragraphs.\\

@@

The Erd\H{o}s-R\'{e}nyi model is used as a benchmark model. In this thesis we focus intensively on two network features/topologies that may impact the formation of echo chambers, namely a high clustering coefficient and community structure. It is therefore necessary to have a model that lacks both of these features to compare the results with. The Erd\H{o}s-R\'{e}nyi model is one of the most simple random network models that does not display both of these features and is hence the perfect comparison model. \\

@@

The Watts-Strogatz model is used in as a general model to investigate the effect of high clustering on the evolution of opinions and the formation of echo chambers.\\

@@

In the paper \textit{``Modelling opinion dynamics in the age of algorithmic personalization''} (reference \cite{Perra2019}) Perra and Rocha showed that a high clustering coefficient enhances the formation of echo chambers \cite{Perra2019}. Here, we will use the Watts-Strogatz model to investigate the interplay between high clustering and the presence of stubborn actors and we will use it as a comparison for networks that posses a community structure without having a high clustering coefficient. We will investigate whether community structure behaves similar to high clustering regarding the appearance of echo chambers.\\

@@

The stochastic Watts-Strogatz block model is used as a model that has both a high clustering and a community structure. We will check if the combination of both features displays any difference in the formation of echo chambers compared to having only a single feature (either community structure or high clustering).\\

@@

Ideally, we would test the opinion dynamics model that was used throughout this thesis on network data directly obtained from social network companies. However, most of these companies are not to keen on sharing any form of data. This makes it very hard to find accurate and complete social media network data. Hence, we chose to test the model on other types of social networks, either on-line or off-line. Some possibilities are any form of friendship networks, e-mail networks, collaboration networks, etc. \\
\newline
One of the decisive properties of the real-world networks that were eventually used, was their size. The main, and actually only, reason for this was the computation time. If the networks would be to large, it would take weeks, sometimes even months before there would be any kind of results.\\
\newline
The size of the network models that were used, was around a thousand nodes. For the real-world networks, we also used networks with a size of magnitude of thousand to ten-thousand. This benefits the computation time; it is, however, unknown how the network size affects the results. It is therefore recommended to test the model on larger network models and on larger real-world networks in future researches on this topic. \\
\newline
We used two real-world networks. The data was obtained from \textit{The index of Complex networks} (ICON) \cite{ICON}. The first real-world network is an on-line friendship network among Finnish users of the social music listening and sharing website Last.fm (data from $2006$). Only the largest connected component of this on-line social network is included \cite{ICON}. This data set was originally used in reference \cite{Toivonen2009}, which describes the data set. This on-line social network has the following properties: number of nodes: $N = 8003$; number of edges: $16824$; modularity: $Q = 0.79$; average clustering coefficient: $\left<cc\right> = 0.31$; average degree: $\left<k\right> = 4.2$. The degree distribution is heterogeneous and is best fitted by a log-normal
\begin{equation}
	f(k) = \frac{1}{k \sigma \sqrt{2\pi}} \exp(-\frac{(\ln(k) - \mu)^2}{2 \sigma^2})\ ,
\end{equation}
with $\mu = 1.0078272$ and $\sigma = 0.95173779$. The degree distribution of the Last.fm social network, together with the log-normal fit, is shown in Fig. \ref{deg_distr_lastfm}.\\
\newline
The second real-world network that was used is the giant component of the on-line, social network of users of the Pretty-Good-Privacy (PGP) algorithm for secure information interchange (data from $2004$) \cite{ICON}. This network was originally used in reference \cite{Boguna2004}, which gives a detailed description of the network data. The main properties of this network are: number of nodes: $N = 10680$; number of edges: $24340$; modularity: $Q = 0.85$; average clustering coefficient: $\left<cc\right> = 0.27$; average degree: $\left<k\right> = 4.56$. The network has a heterogeneous degree distribution, which is best fitted by two power-laws
\begin{equation}
 f(k) = a k^{-\gamma} \ ,
\end{equation}
with $a = 0.39604951$, $\gamma = 1.53568449$ for $k < 20$ and $a = 5.59021251$, $\gamma = 2.4944057$ for $k \geqslant 20$. The degree distribution for the PGP trust network, together with the power-law fits, is shown in Fig. \ref{deg_distr_pgp}.

@@

In Chapter \ref{chap3} we discussed quite a bit of opinion dynamics models and related topics such as algorithmic personalization, activation mechanisms and stubbornness. The goal of this section is to put everything together and summarize the opinion dynamics model of this thesis.\\
\newline
This thesis follows the opinion dynamics model that was designed by Perra and Rocha in their paper \textit{``Modelling opinion dynamics in the age of algorithmic personalization''} \cite{Perra2019}. \\
\newline
We start with a network with $N = 10^3$ users. Initially, each user is given one of the two opinions A or B. Each time step a random fraction equal to $p= 0.1$ of these users becomes active, where $p$ is the activation probability. As discussed in Chapter \ref{chap3}, Section \ref{actMech}, this thesis makes use of a constant activation probability. Every time step, all active users update their opinion synchronously. This updating is done as follows. First, the active user sends his current opinion to all his neighbors (active or not). Then, he updates his own opinion according to the opinions on his time-line. This updating follows the probabilistic majority model, as explained in Chapter \ref{chap3}, Section \ref{probMaj}. If another model is used, eg. the majority model, this will be explicitly stated. If not, one can assume that the probabilistic majority model is implemented. It is important to realize that what an active user sees on his/her time-line is only a selection of all the posts that his/her friends have posted. Each user has a hidden time-line $L_i(t)$. This hidden time-line contains the time-line of the user at the last time he/she was active ($t_{\text{last}}$) plus all the post that his/her friends posted from $t_{\text{last}}$ until the user becomes active again at time $t$. If a user becomes active, the filtering algorithm (REF, REC or PR), discussed in Chapter \ref{chap3}, Section \ref{filter}, selects $S$ items from this hidden time-line and shows them on the actual time-line of the user. We will usually set $S = 20$, which quantifies the maximum length of the actual time-line.\\

@@

If we want to introduce some kind of stubbornness in the model we will mainly use the stubbornness parameter, as was explained in Chapter \ref{chap3}, Section \ref{stubbBin}. Lets briefly summarize this concept of stubbornness. Each node is given a stubbornness parameter $r_i$, with $0 \leqslant r_i \leqslant 1$; $r_i = 0$ means not stubborn and $r_i = 1$ means completely stubborn. For all intermediate values of $r_i$, a random number $f$ between zero and one is drawn; if $f > r_i$ the node can update his opinion, if $f \leqslant r_i$ the node will keep his current opinion.\\

>>>>

On-line social platforms use algorithms to order and filter the posts that appear on the time-line of an individual \cite{Perra2019}. They do this in order to make the experience on their platforms more pleasant, interesting and convenient \cite{Perra2019}. If social media companies would not use any kind of filtering, there would be too much content to handle and process for the social media user \cite{Bozdag2013}. This in turn would make the on-line social experience unpleasant and inconvenient; the user might feel overwhelmed by the number of information he/she is exposed to \cite{Bozdag2013}. The reason behind this is that people are cognitive and time constrained \cite{Perra2019}; hence, there is only a limited amount of information that humans can process \cite{Bozdag2013}.\\
\newline
The downside of such algorithmic filtering is that it fine-tunes the information it shows based on the user. Even two users with the same friends may then be exposed to different content, based on their previous activities and history \cite{Bozdag2013}. This may lead to the formation of echo chambers in which users get `trapped' in content that confirms their own mindset \cite{Bozdag2013}. \\
\newline
It is important to know and be aware of the possible impacts of algorithmic personalization on the formation and evolution of opinions. Perra and Rocha partly investigated the role of such algorithms on the dynamics of opinions in several networks in \cite{Perra2019}. Since filtering algorithms are often not transparently shared by social media companies, they proposed several personalization algorithms based on three simple ways of filtering \cite{Perra2019}: (i) Popularity filtering: promoting content that is popular across the platform; (ii) Semantic filtering: promoting posts that are similar to previous consumed posts; (iii) Collaborative filtering: promoting content that people similar or close to us consume.\\ 
\newline
The following paragraph briefly reviews the filtering algorithms that Perra and Rocha proposed. \\
\newline
At every time $t$ each platform user has a hidden list $L_i(t)$ that contains all the opinions/posts that he/she received from his/her friends \cite{Perra2019}. The filtering algorithm then selects $S$ items from this list and shows them on the user's time-line (this is the fundamental idea behind algorithmic personalization: only showing a fraction of the possible posts \cite{Perra2019}). The following selection methods are used \cite{Perra2019}: (i) the reference filtering algorithm (REF): randomly select $S$ items from the list $L_i(t)$; (ii) the old filtering algorithm (OLD): the $S$ oldest post in $L_i(t)$ are displayed on the time-line of the user; (iii) the recent filtering algorithm (REC): shows the $S$ most recent items from $L_i(t)$; and (iv) the preference filtering algorithm (PR): sorts the hidden list $L_i(t)$ according to the current opinion of the user and selects the first $S$ elements. This filtering algorithm is based on semantic filtering.\\
\newline
Perra and Rocha showed that the REF, OLD and REC filtering algorithms behave similar in many cases and that the PR filtering algorithm often displays different behavior (such as a larger formation of echo chambers compared to the other three filtering algorithms) \cite{Perra2019}. In this thesis the REC and REF filtering algorithms are used as a reference/benchmark algorithm and the PR filtering algorithm is used as the main filtering algorithm.

@@

Most of the simple opinion dynamics models make use of a static network in which users are always active and always exchanging opinions. In this kind of model the opinion dynamics might go as follows: each time step a random user is selected after which his/her opinion is updated according to the opinion dynamics model that is implemented. Another possibility is updating all users at the same time during each time step. \\
\newline
It might, however, be more realistic if nodes/users become active/inactive over time. This would be a better representation of real social media platforms, where users are active for some periods of time and then become inactive again. This kind of activation mechanism of the users gives the network a temporal character. The underlying network structure does not change (each user/node has a fixed number of neighbors that does not change over time, so no new friends are made or, conversely, no old friends are being defriended), but by making users active and inactive the exchange of information between users becomes time dependent and depends on which users are active and which are not. Hence, this does affect the network structure in some way, because if a user, for example, does not become active for a long period of time, he/she does not participate in the exchange of information/opinions and the network structure behaves like if this user/node is absent. This may affect the dynamics, since it may seem as if the network has a different average degree and some properties like connectivity, clustering coefficient, etc may behave differently. \\
\newline
In this thesis we follow the same activation mechanism as was proposed in the paper of Perra and Rocha \cite{Perra2019}. Every time step users are made active with some probability $p_i$ (thus, every time step, each node becomes active with a probability $p_i$). In this thesis the activation probability is set equal to $p = 0.1$, which is constant for all the nodes. Perra and Rocha chose activation probabilities that are extracted from a power-law ($F(p) \sim p^{-1.5}$, with $p \in \interval{0.01}{1}$) \cite{Perra2019}. They did this based on the analyses of real social networks that shows that the propensity of users to start an on-line social interaction follows a heterogeneous distribution \cite{Perra2012}\cite{Perra2019}. However, they also showed that using a constant activation probability $p$ over the entire population yields results that are qualitatively similar (at least in their case) \cite{Perra2019}. \\
\newline
For simplicity, this thesis implements the constant activation probability variant. It might however be an interesting extension to switch to the heterogeneous activation probabilities and investigate the possible impacts on the dynamics. \\
\newline
If a user is active, he/she first spreads his current opinion to all his/her neighbors. Then, the active user updates his own opinion according to the opinions on his/her time-line \cite{Perra2019}. This updating goes according to the implemented opinion dynamics model (eg. majority model, probabilistic majority model,...). In every time step the updating of all active nodes is done simultaneously \cite{Perra2019}.

@@

It is well known that in real life, you can find people that are persistent towards their own opinion and that are not open to opinions of others. They are convinced that they carry the truth with them and that other people, with different world-views, are not capable of giving them any new and useful insights. They are close-minded in contrast to people with an open-mind that are willing to learn from others and to reconsider their beliefs. Of course this is put a bit simplistic and there are many shades of gray between this black and white statement. However, there are groups in society, often with more extreme opinions, that are know to be inflexible towards changing that (extreme) opinion (being stubborn and being extreme does not need to go hand in hand, you can be stubborn without being extreme and vice versa; however, it is often widely accepted that there is some kind of correlation/link between people with extreme opinions and being more persistent towards these opinions compared to people with more moderate beliefs; this idea of correlating extreme people with a stubborn character is often used in models \cite{Deffuant2006}\cite{Deffuant2002}\cite{Turner}) (having some resistance to change your opinion to any other opinion does not necessarily need to be a bad thing, it can be seen as a form of critical thinking and as a form of taking things with a grain of salt).\\ 
\newline
Knowing the impact of stubborn groups on the opinions of others is of great importance, especially with the rise of social media which makes the reach of more extreme and persistent groups a lot bigger. Before the era of social media, one could maybe argue that these extreme and stubborn groups are often isolated and therefore would not reach too many people. So their impact could only be of a minor nature (this is only a statement that could have been given, there are of course plenty of examples throughout human history that may contradict this argument; to only mention a few: Nazism in the 1930s and 1940s, the Ku Klux Klan in the late 19th and begin 20th century, the rise of communism in the 20th century, etc, all examples of groups that can be considered extreme and that have had a (relatively) wide reach and a significant impact on society). Today, this possible argument is no longer valid; social media enables the contact with all sorts of beliefs and all sorts of people with almost no (time and spatial) constraints. Extreme and stubborn groups that may have been isolated before no longer need to be so and this may have huge, and largely unknown, impacts. \\
\newline
What is the impact of having stubborn/extreme people in your society? Does it matter if the stubborn people are grouped together or if they are distributed evenly throughout your system? Are there ways to reduce the possible impacts of stubbornness and/or extremism? Can these stubborn groups lead to polarization or even overtake the opinions in the system? Or is introducing some kind of stubbornness, on the other hand, a protective factor towards extremism/polarization? To answer these questions, several models have been proposed to include some kind of stubbornness/resistance in the system under study.

@@

The bounded confidence models, discussed in Section \ref{bounded}, can be easily extended to included some notion of stubbornness. Guillaume Deffuant proposed such an extension where stubbornness is correlated to extremism \cite{Deffuant2006}. The model he proposed goes as follows. In bounded confidence models, the agents have an opinion $x_i$ (which is a real number between $\interval{-1}{1}$) and a threshold $\epsilon_i$, which determines the range of opinions that can influence agent $i$ (agents with opinions between $\interval{x_i - \epsilon_i}{x_i + \epsilon_i}$ are capable of influencing agent $i$). Guillaume Deffuant then introduced a fraction of extremists, $f_e$, which can have either the extreme opinion $-1$ or the extreme opinion $1$ \cite{Deffuant2006}. These extremists are given a threshold equal to $\epsilon_e$ \cite{Deffuant2006}. The other agents are considered to be moderate and can have opinions between $-1$ and $1$. The moderates are given a threshold $\epsilon$, with $\epsilon > \epsilon_e$ \cite{Deffuant2006}. This last condition implicitly denotes the resistance of the extremes towards other opinions in contrast to the more flexible nature of the moderates, which is based on the idea that people with extreme opinions will be more convinced of that opinion and will be less likely to change it.\\
\newline
Since this thesis is not concerned with bounded confidence models, we will not discuss the results of this model here. Interested readers can find them in \cite{Deffuant2006}. \\
\newline
Another example of introducing some kind of persistence towards the own opinion of the nodes is the ``hardening of positions'' that was briefly discussed at the end of Section \ref{classMod}. The model introduced by Friedkin and Johnsen (Section \ref{classMod}) in which each agent adheres to its initial opinion to a certain degree can also be seen as introducing a notion of stubbornness \cite{Noorazar2020}.

@@

One small remark must be made here: previously, the link between stubbornness and extremism was often made and it was often assumed that the more extreme somebodies opinions are, the more persistent that somebody is towards those opinions. In binary models, however, there are only two possible opinions and in that sense it is less useful to talk about extreme and moderate opinions. However, we can still have people that rather keep their own opinion than be convinced by friends to switch to the other opinion.\\

@@

There are different possibilities of implementing this model. One way, which makes use of the majority model, is as follows. First the fraction of friends with opinion A and opinion B, $p_\text{A}$ and $p_\text{B}$ respectively, are determined; then the following rules are used to update the opinion: if $p_\text{A} > p_\text{B}$ and $p_\text{A} > T$, opinion A is adopted; if $p_\text{B} > p_\text{A}$ and $p_\text{B} > T$, opinion B is adopted; in all other cases, the current opinion of the node is not changed.\\

@@

If $T = 0$, this model reduces to the pure majority model described in Section \ref{majModel}. If $T = 1$, the nodes do not change their opinion. For $0 < T < 1$, the model is a majority model, but the nodes are more persistent towards their own opinion and need a larger fraction of friends to convince them to switch to the other viewpoint.\\

@@

The above examples are only a handful of the possible ways to include some notion of stubbornness in the model, many more exist or can be invented.

@@

The previous two chapters, Chapter \ref{chap2} and \ref{chap3}, gave a broad introduction/overview of networks and opinion dynamics models respectively. In this chapter we will summarize the specific networks and opinion dynamics models that are used in this thesis. \\
\newline
Section \ref{networkmodel} gives an overview of the network models that are used. In chapter \ref{chap2}, Section \ref{netModel} we discussed their main properties; here we will discuss why we use these networks. Which of their properties are of interest in this thesis? What are their benefits and possible shortcomings regarding to our goals?\\
\newline
In section \ref{realnetwork}, we will discuss the used real-world networks. The reasons of choosing these specific networks will be explained, together with their possible strengths and weaknesses.\\ % + explain where they come from etc?
\newline
Finally, Section \ref{opinionmodel} summarizes and discusses the opinion dynamics model that is implemented. We will focus on possible advantages and disadvantages of the model.

@@

This thesis uses three network models: (i) the Erd\H{o}s-R\'{e}nyi model (ER); (ii) the Watts-Strogatz model (WS); and (iii) the stochastic block model, of which two types are implemented: the regular stochastic block model (SBM) and the stochastic Watts-Strogatz block model (SBM-WS).\\

@@

Each of these networks serve specific goals which will be shortly discussed in the following paragraphs.\\

@@

The Erd\H{o}s-R\'{e}nyi model is used as a benchmark model. In this thesis we focus intensively on two network topologies that may impact the formation of echo chambers: a high clustering coefficient and community structure. It is therefore necessary to have a model that lacks both of these features to compare the results with. The Erd\H{o}s-R\'{e}nyi model is one of the most simple random network models that does not display both of these features and is hence the perfect comparison model. \\

@@

The Watts-Strogatz model is used as a general model to investigate the effect of high clustering on the evolution of opinions and the formation of echo chambers.\\

@@

Perra and Rocha showed that a high clustering coefficient enhances the formation of echo chambers \cite{Perra2019}. Here, we will use the Watts-Strogatz model to investigate the interplay between high clustering and the presence of stubborn actors and we will use it as a comparison for networks that posses a community structure without having a high clustering coefficient. We will investigate whether community structure behaves similar to high clustering regarding the appearance of echo chambers.\\

@@

The stochastic Watts-Strogatz block model is used as a model that has both high clustering and community structure. We will check if the combination of both features displays any difference in the formation of echo chambers compared to having only a single feature (either community structure or high clustering).\\

@@

Ideally, we would test the opinion dynamics model that was used throughout this thesis on network data directly obtained from social network companies. However, most of these companies are not to keen on sharing any form of data. This makes it hard to find accurate and complete social media network data. Hence, we chose to test the model on other types of social networks, either on-line or off-line. Some possibilities are any form of friendship networks, e-mail networks, collaboration networks, etc. \\
\newline
One of the decisive properties of the real-world networks that were eventually used, was their size. The main, and actually only, reason for this was the computation time. If the networks are too big, it would take weeks, sometimes even months before there would be results.\\
\newline
The size of the network models that were used, was around a thousand nodes. For the real-world networks, we also used networks with a size of magnitude of thousand to ten-thousand. This benefits the computation time; it is, however, unknown how the network size affects the results. It is therefore recommended to test the model on larger network models and on larger real-world networks in future researches on this topic. \\
\newline
We used two real-world networks. The data was obtained from \textit{The index of Complex networks} (ICON) \cite{ICON}. The first real-world network is an on-line friendship network among Finnish users of the social music listening and sharing website Last.fm (data from $2006$). Only the largest connected component of this on-line social network is included \cite{ICON}. This data set was originally used in \cite{Toivonen2009}, which describes the data set. This on-line social network has the following properties: $N = 8003$; number of edges: $16824$; $Q = 0.79$; $\left<cc\right> = 0.31$; $\left<k\right> = 4.2$. The degree distribution is heterogeneous and is best fitted by a log-normal
\begin{equation}
	f(k) = \frac{1}{k \sigma \sqrt{2\pi}} \exp(-\frac{(\ln(k) - \mu)^2}{2 \sigma^2})\ ,
\end{equation}
with $\mu = 1.0078272$ and $\sigma = 0.95173779$. The degree distribution of the Last.fm social network, together with the log-normal fit, is shown in Fig. \ref{deg_distr_lastfm}.\\
\newline
The second real-world network that was used is the giant component of the on-line, social network of users of the Pretty-Good-Privacy (PGP) algorithm for secure information interchange (data from $2004$) \cite{ICON}. This network was originally used in \cite{Boguna2004}, which gives a detailed description of the network data. The main properties of this network are: $N = 10680$; number of edges: $24340$; $Q = 0.85$; $\left<cc\right> = 0.27$; $\left<k\right> = 4.56$. The network has a heterogeneous degree distribution, which is best fitted by two power-laws
\begin{equation}
 f(k) = a k^{-\gamma} \ ,
\end{equation}
with $a = 0.39604951$, $\gamma = 1.53568449$ for $k < 20$ and $a = 5.59021251$, $\gamma = 2.4944057$ for $k \geqslant 20$. The degree distribution for the PGP trust network, together with the power-law fits, is shown in Fig. \ref{deg_distr_pgp}.

@@

In Chapter \ref{chap3} we discussed opinion dynamics models and related topics such as algorithmic personalization, activation mechanisms and stubbornness. The goal of this section is to put everything together and summarize the opinion dynamics model of this thesis.\\
\newline
This thesis follows the opinion dynamics model that was designed by Perra and Rocha in \cite{Perra2019}. We start with a network with $N = 10^3$ users. Initially, each user is given one of the two opinions A or B. Each time step a random fraction equal to $p= 0.1$ of these users becomes active, where $p$ is the activation probability. As discussed in Chapter \ref{chap3}, Section \ref{actMech}, this thesis makes use of a constant activation probability. Every time step, all active users update their opinion synchronously. This updating is done as follows. First, the active user sends his current opinion to all his neighbors (active or not). Then, he updates his own opinion according to the opinions on his time-line. This updating follows the probabilistic majority model, as explained in Chapter \ref{chap3}, Section \ref{probMaj}. If another model is used, eg. the majority model, this will be explicitly stated. If not, one can assume that the probabilistic majority model is implemented. It is important to realize that what an active user sees on his/her time-line is only a selection of all the posts that his/her friends have posted. Each user has a hidden time-line $L_i(t)$. This hidden time-line contains the time-line of the user at the last time he/she was active ($t_{\text{last}}$) plus all the post that his/her friends posted from $t_{\text{last}}$ until the user becomes active again at time $t$. If a user becomes active, the filtering algorithm (REF, REC or PR), discussed in Chapter \ref{chap3}, Section \ref{filter}, selects $S$ items from this hidden time-line and shows them on the actual time-line of the user. We will usually set $S = 20$, which quantifies the maximum length of the actual time-line.\\

@@

If we want to introduce some kind of stubbornness in the model we will mainly use the stubbornness parameter, as was explained in Chapter \ref{chap3}, Section \ref{stubbBin}. Lets briefly summarize this concept of stubbornness. Each node is given a stubbornness parameter $r_i$, with $0 \leqslant r_i \leqslant 1$; $r_i = 0$ means not stubborn and $r_i = 1$ means completely stubborn. For all intermediate values of $r_i$, a random number $f$ between zero and one is drawn; if $f > r_i$, the node can update his opinion, if $f \leqslant r_i$, the node will keep his current opinion.\\
